Natural variations in water discharge quantity have an inconsistent impact on sediment transport capacity in subglacial channels

### 2023-10-10 Delaneyetal2023b.tex
small bit before mauro's comments

In a transport-limited regime, sediment discharge responds to the sediment transport capacity, defined as the amount of sediment the water can carry.
Sediment transport capacity depends on the shear stress between water and the sediment it flows  over \cite{shields1936,meyer1948,engelund1967} along with the width of the channel bottom $w$ over which to mobilize sediment.
The shear stress  across the channel width $\tau_t$ responds to the velocity of water $v$ flowing through the channel so that
\begin{linenomath*}
  \begin{equation}
    \label{eq:tau_t}
    \tau_t \propto w\, v^2,
  \end{equation}
\end{linenomath*}
%
where $w$ is the width of the channel. Following mass conservation, the velocity of the water flowing through a  channel is
\begin{linenomath*}
  \begin{equation}
    \label{eq:v}
    v = \frac{Q}{S},
  \end{equation}
\end{linenomath*}
where $Q$ is water discharge,  and $S$ is the channel's wetted area.

@@

The response of water velocity to changing water discharge in subglacial channels differs from subaerial ones.
The size of subglacial channels responds to the creep closure of the ice above and the opening of the channel by frictional heating of water flowing through the channel \cite{rothlisberger1972}.
As a result, the subglacial channel size only evolves relatively slowly over days to months, whereas water discharge can vary over hours \cite<e.g.>{iken1986,andrews2014,nanni2020}.
Therefore, subglacial water flow behaves more like pipe flow, and changes in water discharge $Q$ are mainly accommodated with increased or decreased water velocity $v$ \cite{alley1997} because the size of the channel $S$ changes on a much longer timescale (Equation~\ref{eq:v} and Figure~\ref{fig:cartoon}).
%It follows that an increase in water discharge $Q$ will cause the shear stress across the channel $\tau_t$ to respond with water velocity $v$, as opposed to subaerial channels where channel width $w$ evolves as well.

  \begin{figure}[h]
  \centering
    \includegraphics[width=0.8\linewidth]{Fig1.pdf}
    \caption{Sketch for the different responses that subglacial and subaerial channels to increased water discharge.
      Water velocity magnitudes in the subglacial, $v_g$, and subaerial, $v_f$, channels are shown by arrow length.
      $S_g$ and $S_f$ represent the wetted area in subglacial and subaerial channels, respectively.
      The subglacial channel widths $w_g$ remain unchanged, while the subaerial channel widths $w_f$ evolve. Shear stress $\tau$ is responsible for the mobilization of sediment.
      %Note that in the subaerial channel parameterization, channels have rectangular shapes with their width much larger than their depth (Section~\ref{sect:fluv}).
    }
    \label{fig:cartoon}
  \end{figure}
% \ian{Come back to Daniel's comments}

@@

This manuscript evaluates the relationship between sediment transport capacity and water discharge in subglacial channels and compares it to subaerial ones.
In the absence of continuous measurements of subglacial and subaerial water velocity and channel morphology, numerical parameterizations of subglacial and subaerial hydraulics are utilized.
Parameterizations are applied to hydrological records from an Alpine glacier in Switzerland (Fieschergletscher) and  a land-terminating glacier in Greenland (Leverett Glacier).
The parameterizations' lumped nature isolates the disparity between water discharge and sediment transport capacity in subglacial systems, independent of the upstream drainage network and sediment access.
Outputs demonstrate differences in the relationships between  water discharge, channel geometry, and water velocity in the subglacial and subaerial channels.
Furthermore, outputs are compared to sediment discharge data from these two catchments.
The manuscript then uses the demonstrated differences between the two channel types to discuss the implications for interpreting the processes driving variability in records of sediment transport from glacierized catchments.

@@

The Leverett Glacier in Greenland (\icesheet) serves as the ice sheet setting.
Water discharge and suspended sediment concentration data were collected roughly $2$\unit{km} downstream from the terminus between June and August/September 2009 to 2012 ($67^\circ\,03'\,5''$ N, $50^\circ\,12'\,59''$ W).
Data is used from \citeA{tedstone2017} and \citeA{hawkings2014} at a $5$\,\unit{min} time interval from May 28, 2012 to August 8, 2012.
Two gaps (14.07.2012 to 15.07.2012 and 16.07.2012 to 17.07.2021) in suspended sediment data were interpolated using a linear spline.

@@

The width-integrated shear stress is represented as
\begin{equation}
  \label{eq:tautg}
  \tau_{tg}=w_g\,\tau_g,
\end{equation}

where  channel floor width $w_g$ is
\begin{equation}
  \label{eq:dh2wc}
  w_g = 2  \sin \frac{\beta}{2} \sqrt{\frac{2\, S_g}{\beta -\sin \beta}}.
\end{equation}

@@

To parameterize the shear stress of water flowing across sediments in the subaerial channel, the hydraulics parameterization presented in \citeA{tucker1997} is implemented.
Using mass conservation and the Darcy-Weisbach relationship as well as assuming that the channel is wide so that the hydraulic radius is well approximated by the flow depth, then
the shear stress $\tau_f$ at the river bed is
\begin{linenomath*}
  \begin{equation}
    \label{eq:DW_tau}
    \tau_f=\frac{\rho_w\,g^{\frac{2}{3}}\,f_p^{\frac{1}{3}}}{2}\, \Big(\frac{Q}{w_f} \Big)^{\frac{2}{3}} \,\nabla z_c^{\frac{2}{3}},
  \end{equation}
\end{linenomath*}
where $\nabla z_c$ is the channel slope, and $f_p$ is the friction factor for subaerial channels.
Channel width $w_f$ is
\begin{equation}
  \label{eq:wcf}
  w_f = k \, Q^{\alpha},
\end{equation}
%
where $k$ is a constant and $\alpha=\frac{1}{2}$ is a commonly chosen exponent \cite{leopold1953}.
Water velocity, $v_f$, is given by rearranging Equation \ref{eq:tau} as
\begin{equation}
  \label{eq:vf}
  v_f = \sqrt{\frac{8\,\tau_f}{f_p\,\rho_w}}.
\end{equation}
%
As above, the width-integrated shear stress is
\begin{equation}
    \label{eq:tautf}
    \tau_{tf}=w_f\,\tau_f.
  \end{equation}

@@

The parameterizations above are applied to hydrological records from the Fieschergletscher (\alpine) and the Leverett Glacier (\icesheet).
The subglacial and subaerial parameterizations represent the subglacial flow of water exiting a glacier and transitioning to subaerial flow.
Outputs of the parameterizations represent generalizable sediment transport characteristics from these hydrographs, rather than actual hydraulic conditions.
To generalize these cases, \alpine{}  is exemplified by relatively thin ice thickness ($h_{ice}$: $225$\,\unit{m}), low water discharge ($\sim\,10$\,\unit{m}$^3$\,\unit{s}$^{-1}$) and high diurnal variability in water discharge (Figure~\ref{fig:model_outs}, a).
\icesheet{}  is exemplified by thick ice  ($h_{ice}$: $700$\,\unit{m}), high water discharge ($\sim\,300$\,\unit{m}$^3$\,\unit{s}$^{-1}$)  and low diurnal variability in water discharge (Figure~\ref{fig:model_outs}, e).

For each glacier, the parameterizations are first applied to a reference test case for each glacier that assumes  a subglacial channel with $\beta=\frac{\pi}{2}$.
Friction factors $f_r$ and $f_p$ are tuned so that the parameterizations reproduce reasonable water velocities for both \alpine and \icesheet{}\cite<$\sim\,1.6\,$\unit{m}$^3$\,\unit{s}$^{-1}$>{werder2010b,chandler2013}.
In both subaerial channels, the bed slope is $\nabla z_c = 0.05$.
% Water velocity, shear stress, and width-integrated shear stress in the subglacial and subaerial channels from this formulation are presented below.

Second, to further characterize the variability in sediment discharge capacity in subglacial compared to subaerial channels, $100$  parameterizations are applied to a range of different channel geometry factors ($\beta$ and $k$) and friction factors ($f_r$ and $f_p$).
Runs with parameter combinations are culled if their mean subglacial water velocity over the season lies between $0.75$\,\unit{m}$^3$\,\unit{s}$^{-1}$ and $2$\,\unit{m}$^3$\,\unit{s}$^{-1}$ or if subaerial water velocity lies between $0.3$\,\unit{m}$^3$\,\unit{s}$^{-1}$ and $1.2$\,\unit{m}$^3$\,\unit{s}$^{-1}$, in accordance with measurements \cite<e.g.>{werder2010b,magnusson2012,chandler2013}.
The variability of output is represented by the  standard deviation of that output over the course of the season.

Lastly, outputs are compared with the sediment discharge records of the glaciers using Spearman rank correlation, dependent on the ordering of values.
Rank correlation reduces the impact of the non-linear response of sediment transport capacity to the hydrological forcing.
Outputs and observations at $15$\,\unit{min} intervals are aggregated across timescales of up to $15$\,\unit{days} by averaging values one-half of the timescale before and after a point time.

@@

Across the range of channel shape parameters and friction factors  examined, variability in water velocity and shear stress remains higher in the subglacial system compared to the subaerial one.
In \alpine{}, the variability in width-integrated shear stress is comparable between the two systems, due to the variable water discharge (Figure~\ref{fig:multi_run}).
However, in \icesheet{}, with lower water discharge fluctuations, the variability in width-integrated shear stress remains higher in the subglacial system across the time aggregation periods.

@@

First, these measurement stations lie $0.5$ to $2$\,\unit{km} downstream of the glacier terminus \cite{cowton2012,felix2022}, a long enough distance for the pressurized subglacial regime to transition to the open channel subaerial one.
The transition of subglacial to subaerial conditions may occur underneath glacier terminus \cite{perolo2018}, especially if closure rates of the subglacial channels are slow  \cite{rothlisberger1972}.
Furthermore, if a transport-limited regime persists at the glacier margin, then sediment can be easily mobilized, and subaerial processes could be represented just a short distance downstream.

@@

Over time scales of weeks or longer, the glacier's sediment transport capacity responds to the ice thickness, controlling the channel closure rate, and the glacier's surface slope, in addition to water discharge \cite<Figure~\ref{fig:multi_run}, Section~\ref{sect:sub_mode}; >{rothlisberger1972,shreve1972,gimbert2016,stevens2022}.
In a supply-limited state, often assumed in landscape evolution models, then glaciers' sediment discharge record also represents additional processes of sediment availability and bedrock erosion from  water pressure variations and sliding  \cite<e.g.>{iverson2012,herman2015,delaney2023}.
This multitude of processes lies in contrast to most subaerial systems, where transport capacity typically responds to sediment size, water discharge, and hydraulic gradient, and the hydraulic gradient generally remains stable over time scales ranging from years to millennia \cite<Section~\ref{sect:fluv}; e.g.>{muller1968,tucker1997,wickert2019}.
The stronger link between sediment transport capacity and hydrological forcing in subaerial channels may mean that variations in sediment transport records can be more easily attributed to climatic conditions than in subglacial systems.

@@

\appendix
\section{Table of variables, parameters and constants}
\begin{table}[h]
  \centering
  \caption{Variables, parameters, and constants used in this work.
    Where two values are given, the first refers to  \alpine{}, a case from Fieschergletscher, and the second to a glacier marginal to the \icesheet{} a case from Leverett Glacier}
  \begin{tabular}{ l  c  c c }
    Name &Symbol&  Value&Units \\ \hline
    \textbf{Variables}  & & & \\
    Water discharge  & $Q$& & $\mathrm{m^{3}\,s^{-1}}$ \\
    Water velocity (glacier, subaerial)  & $v_g,\,v_{f}$& & $\mathrm{m\,s^{-1}}$ \\
    Channel wetted area (glacier, subaerial) &  $S_g, S_f$& & $\mathrm{m^2}$     \\
    Channel depth (subaerial) & &$H$& $\mathrm{m}$\\
    Hydraulic diameter &$D_h$&&$\mathrm{m}$\\
    Width of channel floor (glacial, subaerial) & $w_g,w_f$&  & $\mathrm{m}$     \\
    Hydraulic head &$\Delta h$&& $\mathrm{m}$\\
    Hydraulic gradient &$\Psi=\frac{\Delta h}{l}$&& $\mathrm{m\, m^{-1}}$\\

@@

\mauro{Notation: either use $h$ and $w$ for the river, or otherwise, somehow free $h$. Below I use $H$ and $W$ for now.}

Here three kinds of drainage regimes are examined, namely, subaerial channels, steady-state R-channels \cite{rothlisberger1972}, and pipe-flow (i.e. R-channels which do not have time to adjust their size to discharge conditions).
Here we show how sediment transport capacity scales with respect to a given water discharge and hydraulic gradient for three different transport formulas: MPM \cite{meyer1948}, EH \cite{engelund1967}, and Bagnold \cite{bagnold1980}; additionally width-integrated shear stress is assessed, as used as proxy for sediment transport throughout the manuscript.

A few basic hydraulic relations are used in the following.
The Darcy-Weisbach equation can be stated as
\begin{equation}
  \label{eq:DW}
  \Psi \propto f\frac{Q^2}{D_h S^2},
\end{equation}
with $\Psi = \Delta h / l$ the head gradient and friction factor $f$.
The shear stress and stream power are, respectively,
\begin{equation}
    \label{eq:tau-omega}
  \tau \propto f v^2 = f \left(\frac{Q}{S}\right)^2, \quad  \Omega \propto \Psi Q.
\end{equation}
%

Sediment discharge is given by the MPM, EH and Bagnold formulations as
\begin{equation}
  Q_s \propto w\, \tau^{3/2}, \quad Q_s \propto w\, \tau^{5/2}, \quad Q_s \propto w \left(\frac{\Omega}{w}\right)^{3/2} H^{-2/3},
\end{equation}
respectively, for conditions well above the transport onset threshold.

The  subaerial channel is assumed to have a width $w$ much greater than its depth $H$, such that $D_h\approx 4h$, and to have a constant head gradient ($\Psi$) given by the topography.
Further, it is  assumed that its width can be approximated by a relation $w \propto Q^\alpha$ (Equation~\ref{eq:wcf})with $\alpha\in [0,1]$; typically $\alpha \approx 1/2$ (a so-called regime channel).
End members $\alpha=0$ or $\alpha=1$ correspond a  subaerial channels of constant width (a slot canyon) or depth (no natural equivalent), respectively.
%
For a steady state R-channel, it is assumed that  $\Psi$ is constant (approximated by the gradient of the Shreve \citeyear{shreve1972} potential \footnote{Note that the R-channel model presented above (Eq.~\eqref{eq:dS_dt}--\eqref{eq:dh2wc}), calculates $\Psi$ from the time evolving $S_g$ (Eq.~\eqref{eq:dS_dt}) via the Darcy-Weisbach equation~\eqref{eq:dh} and thus no Shreve approximation is then needed.}) and that $S$ adjusts such that it is in steady state with $\Psi$ and $Q$.
% Conversely, pipe flow has $S$ fixed and $\Psi$ adjusts to satisfy the given $Q$.
%
Pipe flow like conditions occur when a R-channel is subjected to rapid discharge variations such that the channel cannot adjust its size significantly on the time scale of the discharge variations.
In this case, it is assumed that the cross-sectional area $S$ is fixed, and $\Psi$ adjusts to the specified $Q$.
For both R-channel and pipe flow, it is assumed assume that $D_h \propto S^{1/2}$.

@@

Table~\ref{tab:Qs2} shows how the proxy $w\tau$ as well as $Q_s$ of the MPM, EH and Bagnold sediment transport formulas scale with respect to $Q$, $\Psi$ or $S$ for the three different regimes and for different $\alpha$.
Remarkably, the Bagnold formula has a negative exponent for $f$ in all but the pipe-flow regime, which seems rather unexpected.
The total transport formula EH gives a slightly stronger dependence on all variables as should be expected due to the larger exponent on $\tau$ of $\frac{5}{2}$ versus  $\frac{3}{2}$  for the others.
Albeit, the sediment transport response in pipe-flow for the Bagnold case is close to EH.
Conversely, the sediment transport proxy $w_g\,\tau_{g}, w_f\,\tau_{f}$ used within this publication scales only as $Q^2$ for pipe-flow, whereas the others scale at least with $Q^3$.
The exponent for the width scaling $\alpha$ only impacts relationship between sediment transport and water discharge in the EH relation in any meaningful way.
However, for the value of $\alpha$ around $\frac{1}{2}$ (a value appropriate for most streams), the exponent on $Q$ for EH is only slightly above what the other transport relations give.

The sediment discharge in the steady state R-channel scales very similarly to the  subaerial channel case for all relations and virtually identically for the $\alpha=0.5$ case.  However, note that the head gradients $\Psi$ are likely higher for comparable $Q$ in an ice-sheet marginal or alpine glacier setting than in a  subaerial channel, as is well described in \citeA{alley1997}, and thus transport rates may still be much higher in a steady state R-channel.

Furthermore, R-channels will rarely operate in steady state as variations in discharge, in particular on the diurnal timescale or the timescale of severe rain or melting events, are too fast for such a channel to reach a steady state.
In such cases they operate more like a pipe of fixed cross-section \cite<e.g.>{gimbert2016}.
Table~\ref{tab:Qs2} shows that in such a situation, the sediment transport scales much more severely with discharge, with the exponent on $Q$ being likely between $3$ and $5$, compared to other two regimes when that exponent is at most $1.3$ (Table~\ref{tab:Qs2}).
Thus fluctuations of discharge on short timescales (on the order of a day) have the potential to cause conditions with very high sediment transport capacities.

>>>>

In a transport-limited regime, sediment discharge responds to the sediment transport capacity, defined as the amount of sediment the water can carry.
Sediment transport capacity depends on the shear stress between water and the sediment it flows  over \cite{shields1936,meyer1948,engelund1967} along with the width of the channel bottom $w_c$ over which to mobilize sediment.
The shear stress  across the channel width $\tau_t$ responds to the velocity of water $v$ flowing through the channel so that
\begin{linenomath*}
  \begin{equation}
    \label{eq:tau_t}
    \tau_t \propto w_c\, v^2,
  \end{equation}
\end{linenomath*}
%
where $w_c$ is the width of the channel. Following mass conservation, the velocity of the water flowing through a  channel is
\begin{linenomath*}
  \begin{equation}
    \label{eq:v}
    v = \frac{Q}{S},
  \end{equation}
\end{linenomath*}
where $Q$ is water discharge,  and $S$ is the channel's wetted area.

@@

The response of water velocity to changing water discharge in subglacial channels differs from subaerial ones.
The size of subglacial channels responds to the creep closure of the ice above and the opening of the channel by frictional heating of water flowing through the channel \cite{rothlisberger1972}.
As a result, the subglacial channel size only evolves relatively slowly over days to months, whereas water discharge can vary over hours \cite<e.g.>{iken1986,andrews2014,nanni2020}.
Therefore, subglacial water flow behaves more like pipe flow, and changes in water discharge $Q$ are mainly accommodated with increased or decreased water velocity $v$ \cite{alley1997} because the size of the channel $S$ changes on a much longer timescale (Equation~\ref{eq:v} and Figure~\ref{fig:cartoon}).
%It follows that an increase in water discharge $Q$ will cause the shear stress across the channel $\tau_t$ to respond with water velocity $v$, as opposed to subaerial channels where channel width $w_c$ evolves as well.

  \begin{figure}[h]
  \centering
    \includegraphics[width=0.8\linewidth]{Fig1.pdf}
    \caption{Sketch for the different responses that subglacial and subaerial channels to increased water discharge.
      Water velocity magnitudes in the subglacial, $v_g$, and subaerial, $v_f$, channels are shown by arrow length.
      $S_g$ and $S_f$ represent the wetted area in subglacial and subaerial channels, respectively.
      The subglacial channel widths $w_{cg}$ remain unchanged, while the subaerial channel widths $w_{cf}$ evolve. Shear stress $\tau$ is responsible for the mobilization of sediment.
      %Note that in the subaerial channel parameterization, channels have rectangular shapes with their width much larger than their depth (Section~\ref{sect:fluv}).
    }
    \label{fig:cartoon}
  \end{figure}
% \ian{Come back to Daniel's comments}

@@

This manuscript evaluates the relationship between sediment transport capacity and water discharge in subglacial channels, and compares it to subaerial ones.
In the absence of continuous measurements of subglacial and subaerial water velocity and channel morphology, numerical parameterizations of subglacial and subaerial hydraulics are utilized.
Parameterizations are applied to hydrological records from an Alpine glacier in Switzerland (Fieschergletscher) and  a land-terminating glacier in Greenland (Leverett Glacier).
The parameterizations' lumped nature isolates the disparity between water discharge and sediment transport capacity in subglacial systems, independent of the upstream drainage network and sediment access.
Outputs demonstrate differences in the relationships between  water discharge, channel geometry, and water velocity in the subglacial and subaerial channels.
Furthermore, outputs are compared to sediment discharge data from these two catchments.
The manuscript then uses the demonstrated differences between the two channel types to discuss the implications for interpreting the processes driving variability in records of sediment transport from glacierized catchments.

@@

The Leverett Glacier in Greenland (\icesheet) serves as the ice sheet setting.
Water discharge and suspended sediment concentration data were collected roughly $2$\unit{km} downstream from the terminus between June and August/September 2009 to 2012 ($67^\circ\,03'\,5''$ N, $50^\circ\,12'\,59''$ W).
Data is used from \citeA{tedstone2017} and \citeA{hawkings2014} at a $5$\,\unit{min} time interval from May 28, 2012 to August 8, 2012.
Two gaps (14.07.2012 to 15.07.2012 and 16.07.2012 to 17.07.2021) in suspended sediment data interpolated using a linear spline.

@@

The width-integrated shear stress is represented as
\begin{equation}
  \label{eq:tautg}
  \tau_{tg}=w_{cg}\,\tau_g,
\end{equation}

where  channel floor width $w_{cg}$ is
\begin{equation}
  \label{eq:dh2wc}
  w_{cg} = 2  \sin \frac{\beta}{2} \sqrt{\frac{2\, S_g}{\beta -\sin \beta}}.
\end{equation}

@@

To parameterize the shear stress of water flowing across sediments in the subaerial channel, the hydraulics parameterization presented in \citeA{tucker1997} is implemented.
Using mass conservation and the Darcy-Weisbach relationship as well as assuming that the channel is wide so that the hydraulic radius is well approximated by the flow depth, then
the shear stress $\tau_f$ at the river bed is
\begin{linenomath*}
  \begin{equation}
    \label{eq:DW_tau}
    \tau_f=\frac{\rho_w\,g^{\frac{2}{3}}\,f_p^{\frac{1}{3}}}{2}\, \Big(\frac{Q}{w_{cf}} \Big)^{\frac{2}{3}} \,\nabla z_c^{\frac{2}{3}},
  \end{equation}
\end{linenomath*}
where $\nabla z_c$ is the channel slope, and $f_p$ is the friction factor for subaerial channels.
Channel width $w_{cf}$ is
\begin{equation}
  \label{eq:wcf}
  w_{cf} = k \, Q^{\alpha},
\end{equation}
%
where $k$ is a constant and $\alpha=\frac{1}{2}$ is a commonly chosen exponent \cite{leopold1953}.
Water velocity, $v_f$, is given by rearranging Equation \ref{eq:tau} as
\begin{equation}
  \label{eq:vf}
  v_f = \sqrt{\frac{8\,\tau_f}{f_p\,\rho_w}}.
\end{equation}
%
As above, the width-integrated shear stress is
\begin{equation}
    \label{eq:tautf}
    \tau_{tf}=w_{cf}\,\tau_f.
  \end{equation}

@@

The parameterizations above are applied to hydrological records from the Fieschergletscher (\alpine) and the Leverett Glacier (\icesheet).
The subglacial and subaerial parameterizations represent the subglacial flow of water exiting a glacier and transitioning to subaerial flow.
Outputs of the parameterizations are meant to represent generalizable sediment transport characteristics from these hydrographs, rather than actual hydraulic conditions.
To generalize these cases, \alpine{}  is exemplified by relatively thin ice thickness ($h_{ice}$: $225$\,\unit{m}), low water discharge ($\sim\,10$\,\unit{m}$^3$\,\unit{s}$^{-1}$) and high diurnal variability in water discharge (Figure~\ref{fig:model_outs}, a).
\icesheet{}  is exemplified by thick ice  ($h_{ice}$: $700$\,\unit{m}), high water discharge ($\sim\,300$\,\unit{m}$^3$\,\unit{s}$^{-1}$)  and low diurnal variability in water discharge (Figure~\ref{fig:model_outs}, e).

For each glacier, the parameterizations are first applied to a reference test case for each glacier that assumes  subglacial channel with $\beta=\frac{\pi}{2}$.
Friction factors $f_r$ and $f_p$ are tuned so that the parameterizations reproduce reasonable water velocities for both \alpine and \icesheet{}\cite<$\sim\,1.6\,$\unit{m}$^3$\,\unit{s}$^{-1}$>{werder2010b,chandler2013}.
In both subaerial channels, the bed slope is $\nabla z_c = 0.05$.
% Water velocity, shear stress, and width-integrated shear stress in the subglacial and subaerial channels from this formulation are presented below.

Second, to further characterize the variability in sediment discharge capacity in subglacial compared to subaerial channels, $100$  parameterizations are applied to a range of different channel geometry factors ($\beta$ and $k$) and friction factors ($f_r$ and $f_p$).
Runs with parameter combinations are culled if their mean subglacial water velocity over the season lies between $0.75$\,\unit{m}$^3$\,\unit{s}$^{-1}$ and $2$\,\unit{m}$^3$\,\unit{s}$^{-1}$ or if subaerial water velocity lies between $0.3$\,\unit{m}$^3$\,\unit{s}$^{-1}$ and $1.2$\,\unit{m}$^3$\,\unit{s}$^{-1}$, in accordance with measurements \cite<e.g.>{werder2010b,magnusson2012,chandler2013}.
Variability of an output is represented by the  standard deviation of an output over the course of the season.

Lastly, outputs are compared with the sediment discharge records of the glaciers using Spearman rank correlation, dependent on the ordering of values.
Rank correlation reduces the impact of the non-linear response of sediment transport capacity to the hydrological forcing.
Outputs and observations at $15$\,\unit{min} intervals are aggregated across timescales of up to $15$\,\unit{days} by averaging values one half of the timescale before and after a point time.

@@

Across the range of channel shape parameters and friction factors  examined, variability in water velocity and shear stress remains higher in the subglacial system compared to the subaerial ones.
In \alpine{}, the variability in width-integrated shear stress is comparable between the two systems, due to the variable water discharge (Figure~\ref{fig:multi_run}).
However, in \icesheet{}, with lower water discharge fluctuations, the variability in width-integrated shear stress remains higher in the subglacial system across the time aggregation periods.

@@

First, these measurement stations lie $0.5$ to $2$\,\unit{km} downstream of the glacier terminus \cite{cowton2012,felix2022}, a long enough distance for the pressurized subglacial regime to transition to the open channel subaerial one.
The transition of subglacial to subaerial conditions may even occur underneath glacier terminus \cite{perolo2018}, especially if closure rates of the subglacial channels are slow  \cite{rothlisberger1972}.
Furthermore, if a transport-limited regime persists at the glacier margin, then sediment can be easily mobilized, and subaerial processes could be represented just a short distance downstream.

@@

Over time scales of weeks or longer, the glacier's sediment transport capacity responds to the ice thickness, controlling the channel closure rate, and the glacier's surface slope, in addition to water discharge \cite<Figure~\ref{fig:multi_run}, ~ Section~\ref{sect:sub_mode}; >{rothlisberger1972,shreve1972,gimbert2016,stevens2022}.
In a supply-limited state, often assumed in landscape evolution models, then glaciers' sediment discharge record also represents additional processes of sediment availability and bedrock erosion from  water pressure variations and sliding  \cite<e.g.>{iverson2012,herman2015,delaney2023}.
This multitude of processes lies in contrast to most subaerial systems, where transport capacity typically responds to sediment size, water discharge, and hydraulic gradient, and the hydraulic gradient generally remains stable over time scales ranging from years to millennia \cite<Section~\ref{sect:fluv}; e.g.>{muller1968,tucker1997,wickert2019}.
The stronger link between sediment transport capacity and hydrological forcing in subaerial channels may mean that variations in sediment transport records can be more easily attributed to climatic conditions than in subglacial systems.

@@

\appendix
\section{Table of variables, parameters and constants}
\begin{table}[h]
  \centering
  \caption{Variables, parameters, and constants used in this work.
    Where two values are given, the first refers to  \alpine{}, a case from Fieschergletscher, and the second to a glacier marginal to the \icesheet{} a case from Leverett Glacier}
  \begin{tabular}{ l  c  c c }
    Name &Symbol&  Value&Units \\ \hline
    \textbf{Variables}  & & & \\
    Water discharge  & $Q$& & $\mathrm{m^{3}\,s^{-1}}$ \\
    Water velocity (glacier, subaerial)  & $v_g,\,v_{f}$& & $\mathrm{m\,s^{-1}}$ \\
    Channel wetted area (glacier, subaerial) &  $S_g, S_f$& & $\mathrm{m^2}$     \\
    Channel depth (subaerial) & &$H$& $\mathrm{m}$\\
    Hydraulic diameter &$D_h$&&$\mathrm{m}$\\
    Width of channel floor (glacial, subaerial) & $w_{cg},w_{cf}$&  & $\mathrm{m}$     \\
    Hydraulic head &$\Delta h$&& $\mathrm{m}$\\
    Hydraulic gradient &$\Psi=\frac{\Delta h}{l}$&& $\mathrm{m\, m^{-1}}$\\

@@

\mauro{Notation: either use $h$ and $w_c$ for the river, or otherwise, somehow free $h$. Below I use $H$ and $W$ for now.}

Here three kinds of drainage regimes are examined, namely, subaerial channels, steady-state R-channels \cite{rothlisberger1972}, and pipe-flow (i.e. R-channels which do not have time to adjust their size to discharge conditions).
Sediment transport capacity is scaled with respect to a given water discharge for three different transport formulas: MPM \cite{meyer1948}, EH \cite{engelund1967}, and Bagnold \cite{bagnold1980}; additionally width-integrated shear stress is assessed, as it is  used as proxy for sediment transport throughout the manuscript.

A few basic hydraulic relations are used in the following.
The Darcy-Weisbach equation can be stated as
\begin{equation}
  \label{eq:DW}
  \Psi \propto f\frac{Q^2}{D_h S^2},
\end{equation}
with $\Psi$ the head gradient and friction factor $f$.
The shear stress and stream power are, respectively,
\begin{equation}
    \label{eq:tau-omega}
  \tau \propto f v^2 = f \left(\frac{Q}{S}\right)^2, \quad  \Omega \propto \Psi Q.
\end{equation}
%

Sediment discharge is given by the MPM, EH and Bagnold formulations as, respectively,
\begin{equation}
  Q_s \propto w_c\, \tau^{3/2}, \quad Q_s \propto w_c\, \tau^{5/2}, \quad Q_s \propto w_c \left(\frac{\Omega}{w_c}\right)^{3/2} H^{-2/3}
\end{equation}
for conditions well above the transport onset threshold.

The  subaerial channels is assumed to have a width $w_c$ much greater than its depth $H$, such that $D_h\approx 4h$, and to have a constant head gradient ($\Psi$) given by the topography.
Further, it is  assumed that its width can be approximated by a relation $w_c \propto Q^\alpha$ (Equation~\ref{eq:wcf})with $\alpha\in [0,1]$; typically $\alpha \approx 1/2$ (a so-called regime channel).
End members $\alpha=0$ or $\alpha=1$ correspond a  subaerial channels of constant width (a slot canyon) or depth (no natural equivalent), respectively.
%
For a steady state R-channel, it is assumed that  $\Psi$ is constant (approximated by the gradient of the Shreve \citeyear{shreve1972} potential) and that $S$ adjusts such that it is in steady state with $\Psi$ and $Q$.  %Conversely, pipe flow has $S$ fixed and $\Psi$ adjusts to satisfy the given $Q$.
%
Pipe flow is approximated when a R-channel is subjected to rapid discharge variations such that the channel does not adjust its size significantly on the time scale of the discharge variations.
In this case, it is assumed that the cross-sectional area $S$ is fixed, and $\Psi$ adjusts to the specified $Q$.
For both R-channel and pipe flow, it is assumed assume that $D_h \propto S^{1/2}$.

@@

Table~\ref{tab:Qs2} shows that the proxy $w_c\tau$ as well as the MPM, EH and Bagnold sediment transport formulas give roughly a similar scaling of $Q_s$ on hydraulic gradient $\Psi$ or $S$ and on water discharge $Q$.
Remarkably, the Bagnold formula has a negative exponent for $f$ in all but the pipe-flow regime, which seems rather unexpected.
The total transport formula EH gives a slightly stronger dependence on all variables as should be expected due to the larger exponent on $\tau$ of $\frac{5}{2}$ versus  $\frac{3}{2}$  for the others.
Albeit, the sediment transport response in pipe-flow for the Bagnold case is close to EH.
Conversely, the sediment transport proxy $t_{tg},t_{tf}$ used within this publication scales only as $Q^2$ for pipe-flow, whereas the others scale at least with $Q^3$.
The exponent for the width scaling $\alpha$ only impacts relationship between sediment transport and water discharge in the EH relation in any meaningful way.
However, for the value of $\alpha$ around $\frac{1}{2}$,  suits braided proglacial stream and indeed most  subaerial channels.
Furthermore, when $\alpha=1$, the exponent on $Q$ is only slightly above the other values of $\alpha$.

The sediment discharge in the steady state R-channel scales very similarly to the  subaerial channel case for all relations and virtually identically for the $\alpha=0.5$ case.  However, note that the head gradients $\Psi$ are likely higher for comparable $Q$ in a ice-sheet marginal or alpine glacier setting than in a  subaerial channel, as is well described in \citeA{alley1997}, and thus transport rates may still be much higher in a steady state R-channel.

Furthermore, R-channels will rarely operate in steady state as variations in discharge, in particular on the diurnal timescale or the timescale of severe rain or melting events, are too fast for such a channel to reach a steady state.
In such cases they operate more like a pipe of fixed cross-section \cite<e.g.>{gimbert2016}.
Table~\ref{tab:Qs2} shows that in such a situation, the sediment transport scales much more severely with discharge, with the exponent on $Q$ being likely between $3$ and $5$, compared to other two regimes when that exponent is at most $1.3$ (Table~\ref{tab:Qs2}).
Thus fluctuations of discharge on short timescales (on the order of a day) have the potential to cause very high sediment transport conditions.

\begin{table}
  \caption{Relations for hydraulic variables for the three drainage regimes:  subaerial channels, R-channel and pipe-flow.  Darcy-Weisbach equation is abbreviated with ``D-W'', and stream power with ``Stream p.''. }
  \label{tab:eqs1}
  \small
\begin{tabular}{llllll}
  Regime & Fixed & Determined via D-W
  \makecell{Equation~\ref{eq:DW}} & Additional relations & Shear stress & Stream p.\\
 &  &  &  & \(\tau \propto\) & \(\Omega \propto\)\\
\hline
  Subaerial & \(\Psi\) & \(H \propto f^{1/3}\, Q^{2/3-2\alpha/3} \, \Psi^{-1/3}\) & 
\makecell{\(w_c\,\propto Q^\alpha\) \\ \(S=w_cH\) \\ \(D_h=H\)} & 
\(f^{1/3} Q^{2/3-2\alpha/3}  \Psi^{2/3}\) & \(Q \Psi\)\\
R-ch. & \(\Psi\) & \(S\, \propto f^{2/5}\, Q^{4/5} \, \Psi^{-2/5}\) & \(D_h\propto w_c \propto H \propto S^{1/2}\) & \(f^{1/5} Q^{2/5} \, \Psi^{4/5}\) & \(Q \Psi\)\\
Pipe & \(S\) & \(\Psi \propto f \, Q^2\, S^{-5/2}\) & \(D_h\propto w_c \propto H \propto S^{1/2}\) & \(f Q^2 S^{-2}\) & \(f Q^3 S^{-5/2}\)\\
\end{tabular}
\end{table}
